---
title: "BIS15L_MT2"
output: html_notebook
---
##SHINY
```{r}
install.packages(shiny)
library(shiny)
```
the `ui` function controls the user inputs and the way the app will be displayed (how the app looks)
the `server` function is the part of the app that takes the values and performs calculations
the `run` function combines both to run the app

typing `shinyapp` gives you a skeleton shinyapp
```{r}
library(shiny)

ui <- fluidPage(
  
)

server <- function(input, output, session) {
  
}

shinyApp(ui, server)
```
##SHINY INPUT
```{r}
library(shiny)

ui <- fluidPage(
  selectInput("x",
              "select x variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
  
  selectInput("y",
              "select y variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm")
  
)

server <- function(input, output, session) {
  
}

shinyApp(ui, server)
```
##SERVER SIDE
```{r}
library(shiny)

ui <- fluidPage(
  selectInput("x",
              "select x variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
  
  selectInput("y",
              "select y variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
  plotOutput("plot",
             width = "500px",
             height = "400px")
)

server <- function(input, output, session) {
  
  output$plot <- renderPlot({
    
    ggplot(data = penguins, 
           aes_string(x=input$x, y=input$y, color="species"))+
      geom_point(na.rm = T)
  })
  
}

shinyApp(ui, server)
```
##SHINY APP PRACTICE
```{r}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")%>%
  clean_names()
```

```{r}
homerange%>%
  ggplot(aes(x=locomotion, fill=trophic_guild))+
  geom_bar(position="dodge")
```

```{r}
library(shiny)

ui <- fluidPage(
  
  radioButtons("x",
               "select fill variable",
               choices= c("trophic_guild",
                          "thermoregulation"),
               selected = "trophic_guild"),
  plotOutput("plot")
)

server <- function(input, output, session) {
  
  output$plot <- renderPlot({
    
    ggplot(data=homerange,
           aes_string(x="locomotion", fill=input$x))+
      geom_bar(position="dodge", alpha = 0.8, color = "black")+
      labs (x=NULL, y=NULL, fill="fill variable")
  })
}

shinyApp(ui, server)
```
## Create Base Map
Our goal is to plot the spiders locations from the columns which contain the latitude and longitude. First, we need to get a base map for plotting our points on. We could plot them without a base map, but that wouldn't give us any context as to where they are in space. To get a base map we specify a min and max of each x and y coordinate, and create a bounding box.  

We set the bounding box to a little outside our min and max locations with `f = 0.03`. This gives an extra 3% padding to the bounding box.   

`summary()` gives us our min and max.
```{r}
spiders %>% 
  select(latitude, longitude) %>% 
  summary()
```
Now we set the bounding box. We use the min and max values for latitude and longitude to set the range.  
```{r}
lat <- c(34.67, 41.80)
long <- c(-124.1, -115.5)
bbox <- make_bbox(long, lat, f=0.03)
```
Let's get a base map for our bounding box area. There are several different map [styles](https://docs.stadiamaps.com/themes/)
```{r}
map1 <- get_stadiamap(bbox, maptype="stamen_terrain", zoom = 7)
```
## Adding Points to Base Map
To add our points we only need to specify the x and y location similar to how we made plots in `ggplot.` 
```{r}
ggmap(map1) +
  geom_point(data=spiders, aes(longitude, latitude), size=1, color="red", alpha=0.8)+
  labs(x="Longitude", y="Latitude", title="CA Usofila Records")
```
## Leaflet
If we want to make an interactive map, we can use the `leaflet` package. This allows us to zoom and pan around the map, and we can add popups with information about each point.  

Get the min/max from the data.
```{r}
spiders %>% 
  summarize(
  lon_min=min(longitude, na.rm=T),
  lon_max=max(longitude, na.rm=T),
  lat_min=min(latitude, na.rm=T),
  lat_max=max(latitude, na.rm=T))
```
Add this to four new variables which will be used to create the bounding box in leaflet.
```{r}
lon_min <- -124.13	
lon_max <- -115.49	
lat_min <- 34.67	
lat_max <- 41.8	
```
Build the interactive map using Stadia maps.
```{r}
leaflet(spiders) %>% 
  addProviderTiles(providers$Stadia.StamenTerrain) %>% 
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius=2,
    stroke = FALSE,
    fillOpacity=0.7
  ) %>% 
  addScaleBar(position="bottomleft") %>% 
  fitBounds(lng1=lon_min, lat1=lat_min,
            lng2=lon_max, lat2=lat_max)
```
Build the same interactive map using default OpenStreet maps.
```{r}
leaflet(spiders) %>% 
  addTiles() %>% 
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius=2,
    stroke = FALSE,
    fillOpacity=0.7
  ) %>% 
  addScaleBar(position="bottomleft") %>% 
  fitBounds(lng1=lon_min, lat1=lat_min,
            lng2=lon_max, lat2=lat_max)
```
## Practice
1. Load the `grizzly` data and evaluate its structure.  
```{r message=FALSE, warning=FALSE}
grizzly <- read_csv("data/bear-sightings.csv")
```

```{r}
names(grizzly)
```
2. Use the range of the latitude and longitude to build an appropriate bounding box for your map.  
```{r}
grizzly %>% 
  select(latitude, longitude) %>% 
  summary()
```
```{r}
lat <- c(55.02, 70.37)
long <- c(-131.3, -166.2)
bbox <- make_bbox(long, lat, f=0.05)
```
3. Load a map from `stadiamaps` in a `stamen_terrain` projection and display the map.
```{r}
map1 <- get_stadiamap(bbox, maptype="stamen_terrain", zoom=5)
```

```{r}
ggmap(map1)
```
4. Build a final map that overlays the recorded observations of grizzly bears in Alaska.  
```{r}
ggmap(map1)+
  geom_point(data=grizzly, aes(longitude, latitude), size=0.8)+
  labs(x="Longitude", y="Latitude", title="Ursos horriblus")
```
## Joining Files
Sometimes data we need is stored in a separate file or becomes available later and we need to join it to our existing data in order to work with it. Let's use an easy example to explore some of the different types of joins. This section was based on the excellent example provided by [Holly Emblem](https://hollyemblem.medium.com/joining-data-with-dplyr-in-r-874698eb8898).  

Let's assume we have a database of customers and their purchases.
```{r message=FALSE, warning=FALSE}
table_A <- read_csv("data/table_A.csv")
table_B <- read_csv("data/table_B.csv")
```

```{r}
table_A
```

```{r}
table_B
```

The general syntax for joins is...
```{r}
#join_type(firstTable, secondTable, by=columnTojoinOn)
```
`inner_join`  
Where did customers 3 and 4 go? Since they are not on both tables, they were dropped. So, `inner_join` only keeps the rows that have matching values in both tables; it is very strict about matching.
```{r}
inner_join(table_A, table_B, by="customer_ID")
```
`left_join`
Where did customer 4 go? Notice the missing data for customer 3. So, `left_join` keeps all the rows in the first table and only the matching rows in the second table. If there is no match, it fills in with NA.
```{r}
left_join(table_A, table_B, by="customer_ID")
```
`right_join`
Where did customer 3 go? Notice the missing data for customer 4. `right_join` does the opposite of `left_join`. It keeps all the rows in the second table and only the matching rows in the first table. If there is no match, it fills in with NA.
```{r}
right_join(table_A, table_B, by="customer_ID")
```
`full_join`
`full_join` keeps all the rows in both tables and fills in with NA where there is no match.
```{r}
full_join(table_A, table_B, by="customer_ID")
```
`anti_join`
Provides the rows in the first table for which there are not matching values in the second table.
```{r}
anti_join(table_A, table_B, by="customer_ID") #use to troubleshoot problematic joins
```
## Dates
The `lubridate` package was created to deal with dates of all types. There are many useful functions in `lubridate`, most of which are designed to make working with dates easier. For example, you can easily get the current date and time with `today()` and `now()`.
```{r}
today()
```

```{r}
now()
```
There are many different options for working with dates and date-times in lubridate including `ymd`, `mdy`, `dmy`, `mdy_hms`, `ymd_hms`. The formats below are common examples of how you might see dates in your data.
```{r}
dateformat1 <- "20200922"
dateformat2 <- "09-22-2020"
dateformat3 <- "22/09/2020"
dateformat4 <- "09-22-2020 17:00:00"
dateformat5 <- "20200922 170000"
```
## Converting dates 
We can convert each of the examples above using lubridate.
```{r}
ymd(dateformat1)
```

```{r}
mdy(dateformat2)
```

```{r}
dmy(dateformat3)
```

```{r}
mdy_hms(dateformat4)
```

```{r}
ymd_hms(dateformat5)
```
## Practice
For this practice, we will use data from my research on _Usofila_ spiders. These spiders live mostly in caves and the data were collected by building a database of all the known museum specimens in the US. There are two files: 1. spider_records.csv which contains all the records of spiders and 2. spiders_locations.csv which contains the lat/long for each record. An accession number is a unique identifier for each specimen.

1. Load the two files as separate dataframes. Make sure to clean the column names.
```{r message=FALSE, warning=FALSE}
spider_records <- read_csv("data/spider_records.csv") %>% clean_names()
```

```{r, message = FALSE}
spider_locs <- read_csv("data/spiders_locations.csv") %>% clean_names()
```
2. Join the files using an appropriate key and join type.
```{r}
spiders_with_locs <- 
  left_join(spider_records, spider_locs, by="accession")
```

```{r}
names(spiders_with_locs) #check to see if it worked
```
3. Convert the date column to a date format using lubridate.
```{r}
spiders_with_locs <-spiders_with_locs %>% 
  mutate(date=dmy(date))
```
4. Save your final data to a `.csv`.
```{r}
write.csv(spiders_with_locs, file="spiders_with_locs.csv", row.names=FALSE)
```
## Review
When working with "wild" data, dealing with NA's is a fundamental part of the data cleaning process. Data scientists spend much of their time cleaning and transforming data- including managing NA's. There isn't a single approach that will always work so you need to be careful about using replacement strategies across an entire data set. And, as the data sets become larger NA's can become trickier to deal with.  

## Load the mammals life history data and clean the names  
```{r message=FALSE, warning=FALSE}
life_history <- read_csv("data/mammal_lifehistories_v3.csv") %>% clean_names()
```
## Are there any NA's?
Sometimes using one or more of the summary functions can give us clues to how the authors have represented missing data. This doesn't always work, but it is a good place to start.  
```{r}
glimpse(life_history)
```

```{r}
summary(life_history)
```
## Where are the NA's?
This will give you a quick summary of the number of NA's in each variable. Notice that, at least for now, it doesn't look like there are NA's in most variables because they are still represented by -999.
```{r}
life_history %>% 
  summarize(across(everything(), ~ sum(is.na(.))))
```
Here we look for all of the -999's.
```{r}
life_history %>% 
  summarize(across(everything(), ~sum(.==-999)))
```
## Dealing with NA's using `read_csv()`
In the case of `-999` as NA, we can use `read_csv` to replace these values when we import the data. But, I only do this once I understand how NA's are represented in the data. Sometimes, they are represented in multiple ways.
```{r message=FALSE, warning=FALSE}
life_history_no_nas <- 
  read_csv("data/mammal_lifehistories_v3.csv", na="-999") %>% 
  clean_names()
```
Rechecking for NA's. Now we see that there are NA's in many of the variables.
```{r}
life_history_no_nas %>% 
  summarize(across(everything(), ~ sum(is.na(.))))
```
## Did we catch them all?
Sometimes it can be helpful to do a quick scan using view() or glimpse() to see if there are any other odd values that might represent NA's. Notice that `not measured` is used in `newborn`.
```{r}
glimpse(life_history_no_nas)
```

```{r}
summary(life_history_no_nas)
```
Notice that `max_life` has no NA's. Does that make sense? How likely is it that we know the lifespan for all of the species in the data set?
```{r}
hist(life_history_no_nas$max_life)
```

```{r}
life_history_no_nas %>% 
  filter(max_life==0) %>% 
  select(order, family, genus, species, max_life)
```
Let's use `mutate()` and use `na_if()` to replace 0's with NA's in `max_life`. This chunk allows us to address problems in a single variable.
```{r}
life_history_no_nas <- life_history_no_nas %>% 
  mutate(max_life=na_if(max_life, 0))
```
## `naniar`
[naniar](https://naniar.njtierney.com/) is a package that manages NA's. Many of the functions it performs can also be performed using tidyverse functions, but it provides some nice alternatives.  

`miss_var_summary` provides a summary of NA's across the data frame.
```{r}
miss_var_summary(life_history_no_nas)
```
A unique feature of `naniar` is that it can produce visuals to help evaluate NA's. Here we use `gg_miss_var` to visualize the number of NA's in each variable.
```{r}
life_history_no_nas %>% 
  gg_miss_var()
```
We can also use `geom_miss_point()` to visualize where NA's are located in a scatter plot. Here we see that there are NA's in `wean_mass` across the range of `mass`.
```{r}
life_history_no_nas %>% 
  ggplot(aes(x=log10(mass), y=log10(wean_mass)))+
  geom_point()+
  geom_miss_point()
```
We can also use `miss_var_summary` with `group_by()`. This helps us better evaluate where NA's are in the data.
```{r}
life_history_no_nas %>% 
  group_by(order) %>% 
  miss_var_summary(order=T)
```
`naniar` has nice replace functions which will allow you to precisely control which values you want replaced with NA's in each variable. This is a nice alternative to `mutate()` and `na_if()`.  
```{r}
life_history %>% #going back to the original data
  replace_with_na(replace = list(newborn = "not measured", 
                                 weaning= -999, 
                                 wean_mass= -999, 
                                 afr= -999, 
                                 max_life= 0, 
                                 litter_size= -999, 
                                 gestation= -999, 
                                 mass= -999)) %>% 
miss_var_summary()
```
You can also use naniar to replace a specific value (like -999) with NA across the entire data set.  
```{r}
life_history %>% 
  replace_with_na_all(condition = ~.x ==-999) %>% 
  miss_var_summary()
```
Finally, naniar has some built-in examples of common values or character strings used to represent NA's. The chunk below will use these built-in parameters to replace NA's across the entire data set.
```{r}
common_na_strings
```

```{r}
common_na_numbers
```

```{r}
life_history %>% 
  replace_with_na_all(condition = ~.x %in% c(common_na_numbers, common_na_strings)) %>% 
  mutate(newborn=na_if(newborn, "not measured"))
```
## Practice
Let's practice evaluating NA's in a large data set. The data are compiled from [CITES](https://cites.org/eng). This is the international organization that tracks trade in endangered wildlife. You can find information about the data [here](https://www.kaggle.com/cites/cites-wildlife-trade-database).  

Some key information:  
[country codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2)  

1. Import the data and do a little exploration. Be sure to clean the names if necessary.
```{r message=FALSE, warning=FALSE}
cites <- read_csv("data/cites.csv") %>% 
  clean_names()
```

```{r}
glimpse(cites)
```
2. Use `naniar` to summarize the NA's in each variable.
```{r}
miss_var_summary(cites)
```
3. Try using `group_by()` with `naniar`. Look specifically at `class` and `exporter_reported_quantity`. For which taxonomic classes do we have the highest number of missing export data?
```{r}
cites %>% 
  select(class, exporter_reported_quantity) %>% 
  group_by(class) %>% 
  miss_var_summary() %>% 
  arrange(desc(n_miss))
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
```

```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```
## ggplot themes
There are many options to change the theme of your plots within ggplot. Have a look [here]https://ggplot2.tidyverse.org/reference/ggtheme.html) for a list of the themes.  

Let's start by building a simple barplot.  
```{r}
p <- homerange %>% 
  ggplot(aes(x=taxon, fill=thermoregulation))+
  geom_bar(na.rm=T, position="dodge")+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
p
```
Have a look at the `linedraw` theme; I am adding it as another layer.
```{r}
p + theme_gray()+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```
## Practice
1. Build a scatterplot that compares `log10_mass` and `log10_preymass`. Color by `thermoregulation.` Store this plot as object `q`.
```{r}
q <- homerange %>% 
  drop_na(log10_mass, log10_preymass) %>% 
  ggplot(aes(x=log10_mass, y=log10_preymass, color=thermoregulation))+
  geom_point()+
  geom_smooth(method="lm", se=F, color="red")
q
```
2. Use plot `q` but add the classic theme.
```{r}
q+theme_classic()
```
## Legends
There are lots of options to manipulate legends. Have a look [here](http://www.sthda.com/english/wiki/ggplot2-legend-easy-steps-to-change-the-position-and-the-appearance-of-a-graph-legend-in-r-software).
```{r}
p+theme_linedraw()+
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

```{r}
p + 
  theme_wsj()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
## Practice
1. Use plot `q` and try a theme of your choice.
```{r}
q+theme_stata()
```
## RColorBrewer
The default colors used by ggplot are uninspiring. They don't make plots pop out in presentations or publications, and you may want to use a customized palette to make things visually consistent.  

Access the help for `RColorBrewer`.
```{r}
?RColorBrewer
```
The thing to notice is that there are three different color palettes: 1) sequential, 2) diverging, and 3) qualitative. Within each of these there are several selections. You can bring up the colors by using `display.brewer.pal()`. Specify the number of colors that you want and the palette name.
```{r}
display.brewer.pal(8,"Accent") #qualitative palette
```
The [R Color Brewer](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) website is very helpful for getting an idea of the color palettes. To make things easy, use these two guidelines:

+`scale_colour_brewer()` is for points  
+`scale_fill_brewer()` is for fills  

Here I chose the `Set2` palette. Take a moment and experiment with other options.
```{r}
p+scale_fill_brewer(palette = "Accent")+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
## Practice
1. Keep using the `same `q` plot, but test out one of R Color Brewer palettes.
```{r}
q+theme_classic()+
  scale_colour_brewer(palette="Accent")
```
## Paleteer
You can also use `paleteer` to build a custom palette. To access the `paleteer` collection, I add it to a new object.
```{r}
colors <- paletteer::palettes_d_names
```
Now we can display the palettes. Assign the palette to `my_palette` and then build this base R bar plot. There are a lot of options; `paleteer` is a collection of popular palettes. I really like the [`ggsci` package] (https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html)
```{r}
my_palette <- paletteer_d("ggsci::uniform_startrek")
my_palette
```

```{r}
barplot(rep(1,7), axes=FALSE, col=my_palette)
```
Now we just identify `my_palette` as part of `scale_fill_manual()`
```{r}
p+scale_fill_manual(values=my_cols)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
You can use custom colors by specifying the hex code. You can find hex codes for colors using [this website](https://r-charts.com/colors/).
```{r}
library(scales)

my_cols <- c("#66CDAA",
             "#FFCD00FF")

show_col(my_cols)
```
## Practice
1. Keep using the `same `q` plot, but test out one of `paleteer` themes.
```{r}
q+theme_classic()+
  scale_colour_manual(values=my_cols)
```
## Faceting
[Faceting](https://ggplot2-book.org/facet.html) is one of the amazing features of ggplot. It allows us to make multi-panel plots for easy comparison.  Here is a boxplot that shows the range of log10_mass by taxon, filled by thermoregulation.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10_mass, fill=thermoregulation))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```
`facet_wrap()` makes a ribbon of panels by a specified categorical variable and allows you to control how you want them arranged.  
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10_mass, fill=thermoregulation))+
  geom_boxplot()+
  facet_wrap(~trophic_guild, nrow=2, labeller=label_both)+
  theme(axis.text.x=element_text(angle=60, hjust=1),
        legend.position="none",
        strip.text=element_text(size=12, face="bold"))
  
```
`facet_grid()` allows control over the faceted variable; it can be arranged in rows or columns. rows~columns. 
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10_mass, fill=thermoregulation))+
  geom_boxplot()+
  facet_grid(.~thermoregulation, labeller=label_both)+
  theme(axis.text.x=element_text(angle=60, hjust=1),
        legend.position="none",
        strip.text=element_text(size=12, face="bold"))
```
`facet_grid()` will also allow the comparison of two categorical variables, just remember a~b where a is rows and b is columns.  
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10_mass, fill=thermoregulation))+
  geom_boxplot()+
  facet_grid(trophic_guild~thermoregulation, labeller=label_both)+
  theme(axis.text.x=element_text(angle=60, hjust=1),
        legend.position="none",
        strip.text=element_text(size=8, face="bold"))
```
## Practice
1. Use facet_wrap to compare the distribution of `log10_mass` by `thermoregulation`.
```{r}
homerange %>% 
  ggplot(aes(x=log10_mass))+
  geom_density(fill="steelblue", alpha=0.3)+
  facet_wrap(~thermoregulation, ncol=2)
```
2. Use facet_grid to compare the distribution of `log10_hra` by `locomotion` and `thermoregulation`.
```{r}
homerange %>% 
  ggplot(aes(x=log10_hra))+
  geom_density(fill="steelblue", alpha=0.3)+
  facet_grid(thermoregulation~locomotion)
```
## Data
For these exercises, we are going to use [Gapminder](https://www.gapminder.org/data/) data on population (pop), GDP per capita (gdp), and life expectancy (lex). These data were downloaded and compiled using the `gapminder_compile.Rmd` script in the data folder.
```{r message=FALSE, warning=FALSE}
gap <- read_csv("data/gapminder.csv") %>% clean_names()
```
## Line plots
Line plots are great when you need to show changes over time. Let's see if we can find downturns in the US economy between 2000-2025.  

Let's start by making a clear x and y so we know what we are going to plot.
```{r}
gap %>% 
  filter((year>=2000 & year<=2025) & country=="USA") %>% 
  ggplot(aes(x=year, y=gdp))+
  geom_line()+
  geom_point(shape=5)+
  scale_x_continuous(breaks=2000:2025, minor_breaks = NULL)+
  theme(axis.text.x = element_text(angle=60, hjust=1))+
  labs(title="GDP per Capita in the USA, 2000-2025",
       x="Year",
       y="GDP")
```

```{r}
glimpse(gap)
```

```{r}
gap %>% 
  filter((year>=2000 & year<=2025) & country=="USA") %>% 
  ggplot(aes(x=year, y=gdp))+
  geom_line(color="red")+
  geom_point(shape=5)+
  scale_x_continuous(breaks=seq(2000, 2025, by=5))+
  theme(axis.text.x = element_text(angle=60, hjust=1))+
  labs(title="GDP per Capita in the USA, 2000-2025",
       x="Year",
       y="GDP")
```
## Histograms
Histograms show the distribution of continuous variables. As students, you have seen histograms of grade distributions. A histogram `bins` the data and you specify the number of bins that encompass a range of observations. For something like grades, this is easy because the number of bins corresponds to the grades A-F. By default, R uses a formula to calculate the number of bins but some adjustment may be required.  

What was the distribution of life expectancy in 1935?
```{r}
gap %>% 
  filter(year==1935) %>% 
  ggplot(aes(x=lex))+
  geom_histogram()
```
Let's rebuild the histogram, but this time we will specify the color and fill. Do a little experimentation on your own with the different colors.
```{r}
gap %>% 
  filter(year==1935) %>% 
  ggplot(aes(x=lex))+
  geom_histogram(color="black", fill="turquoise4", bins=30)+
  labs(title="Distribution of Life Expectancy")
```
## Density plots  
Density plots are similar to histograms but they use a smoothing function to make the distribution more even and clean looking. They do not use bins.
```{r}
gap %>% 
  filter(year==1935) %>% 
  ggplot(aes(x=lex))+
  geom_density(color="turquoise4", fill="turquoise4", alpha=0.4)+
  labs(title="Distribution of Life Expectancy")
```
I like to see both the histogram and the density curve so I often plot them together. Note that I assign the density plot a different color.
```{r}
gap %>% 
  filter(year==1935) %>% 
  ggplot(aes(x=lex))+
  geom_histogram(aes(y=after_stat(density)), color="black", fill="turquoise4")+
  geom_density(color="red")
```
## Practice
1. Antibiotics became widely available in the 1940's. How did life expectancy change between 1930 and 1950? Let's plot the density curves for both years on the same plot. Use the `fill` aesthetic to differentiate the two years.
```{r warning=FALSE}
gap %>% 
  filter(year %in% c(1930, 1950)) %>% #pull out the years
  mutate(year=factor(year)) %>% #change to a factor
  ggplot(aes(x=lex, color=year, fill=year))+
  geom_density(alpha=0.3)+
  labs(title="Distribution of Life Expectancy, 1930 and 1950",
       x="Life Expectancy (years)",
       y=NULL,
       color="Year",
       fill="Year")
```
## Create Categories with mutate and case_when()
`case_when()` is a very handy function from `dplyr` which allows us to calculate a new variable from other variables. We use `case_when()` within `mutate()` to do this.`case_when()` allows us to specify multiple conditions.  

The approximate World Bank Income Categories are:  
- Low income: $1,100 or less
- Lower middle income: $1,100 to $4,500
- Upper middle income: $4,500 to $13,900
- High income: $13,900 or more

Let's use `case_when()` to create a new column `income_category` that categorizes countries into these income categories based on their GDP per capita in 2000.
```{r}
income_cat2000 <- gap %>% 
  filter(year==2000) %>% 
  mutate(income_category=case_when(gdp<=1100 ~ "low",
                                   gdp>1100 & gdp <=4500 ~ "low_middle",
                                   gdp>4500 & gdp<=13900 ~ "upper_middle",
                                   gdp>13900 ~ "high")) %>% 
  select(country, gdp, income_category) %>% 
  mutate(income_category=factor(income_category,
                                levels=c("low",
                                         "low_middle",
                                         "upper_middle",
                                         "high")))
```
## Practice
1. How many countries are in each of the World Bank Income Categories as of the year 2000?
```{r}
income_cat2000 %>% 
  group_by(income_category) %>% 
  summarize(count=n())
```
2. Make a bar plot showing the number of countries in each income category in the year 2000.
```{r}
income_cat2000 %>% 
  drop_na(income_category) %>% #get rid of NA's
  ggplot(aes(x=income_category, fill=income_category))+
  geom_bar()+
  labs(title="Countries in WB Income Categories",
       x="Income Category",
       y=NULL)+
  theme(legend.position="none") #remove the legend
```

```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```
## Data
**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  
```{r message=FALSE, warning=FALSE}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv", 
                      na = c("", "NA", "\\")) %>% 
  clean_names()
```
## Aesthetics
There are many options to create nice plots in `ggplot`. One useful trick is to store the plot as a new object and then experiment with geom's and aesthetics. Let's setup a plot that compares `log10_mass` and `log10_hra.` Notice that we are not specifying a geom.
```{r}
p <- homerange %>% 
  ggplot(aes(x=log10_mass, y=log10_hra))
```
Play with point size by adjusting the `size` argument.
```{r}
p+geom_jitter(size=3)
```
We can color the points by a categorical variable.
```{r}
p+geom_point(aes(color=thermoregulation), size=2)
```
We can also map shapes to another categorical variable.
```{r}
p+geom_point(aes(shape=thermoregulation, color=thermoregulation), size=2)
```
What about just filling in with a color preference or size?
```{r}
p+geom_point(size=3, color="black", alpha=0.6, shape=6, fill="lemonchiffon")
```

```{r}
colors()
```

```{r}
# Plot all available point shapes
df <- data.frame(x = 1:25, y = rep(1, 25), shape = 0:24)

ggplot(df, aes(x, y)) +
  geom_point(aes(shape = factor(shape)), size = 5, fill = "lightblue") +
  scale_shape_manual(values = 0:24) +
  theme_void() +
  theme(legend.position = "none") +
  geom_text(aes(label = shape), vjust = -1, size = 5)
```
## Barplots and multiple variables

Let's start by counting how many observations we have in each taxonomic group.
```{r}
homerange %>% 
  count(taxon, sort=T)
```
Now let's make a bar plot of these data.
```{r}
homerange %>% 
  ggplot(aes(x=taxon))+
  geom_bar()+
  coord_flip()+
  labs(title="Observations by Taxon",
       x="Taxonomic Group",
       y=NULL) #take away the y axis label
```
By specifying `fill=trophic.guild` we build a stacked bar plot that shows the proportion of a given taxonomic group that is an herbivore or carnivore.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic_guild))+
  geom_bar()+
  coord_flip()+
  labs(title="Observations by Taxon",
       x="Taxonomic Group",
       y=NULL) #take away the y axis label
```
We can also have counts of each trophic guild within taxonomic group shown side-by-side by specifying `position="dodge"`.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic_guild))+
  geom_bar(position="dodge")+
  coord_flip()+
  labs(title="Observations by Taxon",
       x="Taxonomic Group",
       y=NULL) #take away the y axis label
```
Here is the same plot oriented vertically.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic_guild))+
  geom_bar(position="dodge")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  labs(title="Observations by Taxon",
       x="Taxonomic Group",
       y=NULL) #take away the y axis label
```
We can also scale all bars to a percentage.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic_guild))+
  geom_bar(position=position_fill())+
  scale_y_continuous(labels=scales::percent)+
  coord_flip()
```
## Practice
1. Make a bar plot that shows counts of individuals by island. Fill by species, and try both a stacked bar plot and `position="dodge"`.
```{r}
penguins %>% 
  ggplot(aes(x=island, fill=species))+
  geom_bar()
```

```{r}
penguins %>% 
  ggplot(aes(x=island, fill=species))+
  geom_bar(position="dodge") #side by side
```
2. Make another bar plot that shows the number of individuals by sex on each island?
```{r}
penguins %>% 
  na.omit() %>% #get rid of NA's
  ggplot(aes(x=island, fill=sex))+
  geom_bar(position="dodge")
```
## Using `group`
In addition to `fill`, `group` is an aesthetic that accomplishes the same function but does not add color. It groups data for plotting.

Here is a box plot that shows `log10.mass` by taxonomic `class.`
```{r}
homerange %>% 
  ggplot(aes(x=class, y=log10_mass))+
  geom_boxplot()
```
I use `group` to make individual box plots for each taxon.
```{r}
homerange %>% 
  ggplot(aes(x=class, y=log10_mass, group=taxon))+
  geom_boxplot()
```
I can also use `fill` to associate the different taxa with a color coded key.
```{r}
homerange %>% 
  ggplot(aes(x=class, y=log10_mass, fill=taxon))+
  geom_boxplot()
```
## Bar Plots
Recall that `geom_bar` is used when you want to count the number of observations in a categorical variable. `geom_col` is used when you have already counted the number of observations (or you have a defined x and y) and want to plot those counts.

Make two bar plots showing the number of observations for each order of mammals using both geom types.  

`geom_bar`
```{r}
life_history %>% 
  ggplot(aes(x=order))+
  geom_bar()+
  coord_flip()
```
`geom_col`
```{r}
life_history %>% 
  count(order, sort=T)
```

```{r}
life_history %>% 
  count(order, sort=T) %>% 
  ggplot(aes(x=order, y=n))+
  geom_col()+
  coord_flip()
```
What if we wanted a bar plot of the mean mass for each order?
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass=mean(mass, na.rm=T)) %>% 
  arrange(desc(mean_mass))
```

```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass=mean(mass, na.rm=T)) %>% 
  arrange(desc(mean_mass)) %>% 
  ggplot(aes(x=order, y=mean_mass))+
  geom_col()+
  coord_flip()
```
There are a few problems here. First, the y-axis is in scientific notation. We can fix this by adjusting the options for the session.
```{r}
options(scipen=999) #cancels scientific notation for the session
```
Next, the y-axis is not on a log scale. We can fix this by adding `scale_y_log10()`.
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass=mean(mass, na.rm=T)) %>% 
  arrange(desc(mean_mass)) %>% 
  ggplot(aes(x=order, y=mean_mass))+
  geom_col()+
  coord_flip()+
  scale_y_log10()
```
Lastly, we can adjust the x-axis labels to make them more readable. We do this using `reorder`.
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass=mean(mass, na.rm=T)) %>% 
  arrange(desc(mean_mass)) %>% 
  ggplot(aes(x=reorder(order, mean_mass), y=mean_mass))+
  geom_col()+
  coord_flip()+
  scale_y_log10()
```
## Scatterplots
Scatter plots allow for comparisons of two continuous variables. Make a scatterplot that compares gestation time and wean mass.
```{r}
life_history %>% 
  ggplot(aes(x=gestation, y=wean_mass))+
  geom_point(na.rm=T)+
  scale_y_log10()+
  geom_smooth(method=lm, se=F, na.rm=T)
```
## Boxplots
Box plots are used to visualize a range of values. So, on the x-axis we have a categorical variable and the y-axis is the range. Make a box plot that compares mass across orders.
```{r}
life_history %>% 
  ggplot(aes(x=order, y=log10(mass)))+ #another way of scaling
  geom_boxplot(na.rm=T)+
  coord_flip()
```
## Aesthetics: Labels
Now that we have practiced scatter plots, bar plots, and box plots we need to learn how to adjust their appearance to suit our needs. Let's start with labeling x and y axes.  

```{r}
names(life_history)
```
Is there a relationship between mass and litter size; i.e. do larger mammals have more offspring?
```{r}
life_history %>% 
  ggplot(aes(x=mass, y=litter_size))+
  geom_point(na.rm=T)+
  scale_x_log10()+
  geom_smooth(method=lm, se=F, na.rm=T)
```
The plot looks clean, but it is incomplete. A reader unfamiliar with the data might have a difficult time interpreting the labels. To add custom labels, we use the `labs` command.
```{r}
life_history %>% 
  ggplot(aes(x=mass, y=litter_size))+
  geom_point(na.rm=T)+
  scale_x_log10()+
  geom_smooth(method=lm, se=F, na.rm=T)+
  labs(title="Mass vs. Litter Size",
       x="Mass (g)",
       y="Litter Size")
```
We can adjust the plot further by specifying the size and face of the text. We do this using `theme()`. The `rel()` option changes the relative size of the title to keep things consistent. Adding `hjust` controls the title position.
```{r}
life_history %>% 
  ggplot(aes(x=mass, y=litter_size))+
  geom_point(na.rm=T)+
  scale_x_log10()+
  geom_smooth(method=lm, se=F, na.rm=T)+
  labs(title="Mass vs. Litter Size",
       x="Mass (g)",
       y="Litter Size")+
  theme(plot.title=element_text(size=rel(2), hjust=.5))
```
## `pivot_longer()`
Recall that we use `pivot_longer()` when our column names actually represent variables. A classic example would be that the column names represent observations of a variable.
```{r}
datasets::USPersonalExpenditure
?USPersonalExpenditure
```
Here we add a new column of expenditure types, which are stored as rownames above, with `mutate()`. The `USPersonalExpenditures` data also needs to be converted to a data frame before we can use the tidyverse functions, because it comes as a matrix.
```{r}
expenditures <- as_tibble(USPersonalExpenditure) %>% 
  mutate(expenditure = rownames(USPersonalExpenditure)) %>% 
  relocate(expenditure)
expenditures
```
## Practice
1. Are these data tidy? Please use `pivot_longer()` to tidy the data.
```{r}
expenditures %>% 
  pivot_longer(-expenditure,
               names_to = "year",
               values_to = "bn_dollars")
```
2. Restrict the data to medical and health expenditures only. Sort in ascending order.
```{r}
expenditures %>% 
  pivot_longer(-expenditure,
               names_to = "year",
               values_to = "bn_dollars") %>% 
  filter(expenditure=="Medical and Health") %>% 
  arrange(bn_dollars)
```
## `separate()`
In this new heart rate example, we have the sex of each patient included with their name. Are these data tidy? No, there is more than one value per cell in the patient column and the columns a, b, c, d once again represent values.
```{r}
heartrate2 <- read_csv("data/heartrate2.csv")
heartrate2
```
We need to start by separating the patient names from their sexes. `separate()` needs to know which column you want to split, the names of the new columns, and what to look for in terms of breaks in the data.
```{r}
heartrate2 %>% 
  separate(patient, into=c("patient", "gender"), sep="_")
```
Another way of using separate is to specify the position of the split. In this case, we want to split the patient column at the last character.
```{r}
heartrate2 %>% 
  separate(patient, into=c("patient", "gender"), sep=-1)
```
We would still need to deal with the "_" in the patient column. We can do this using the stringr package.
```{r}
heartrate2 %>% 
  separate(patient, into=c("patient", "gender"), sep=-1) %>% 
  mutate(patient=str_remove(patient, "_"))
```
## Practice
1. Re-examine `heartrate2`. Use `separate()` for the sexes, `pivot_longer()` to tidy, and `arrange()` to organize by patient and drug. Store this as a new object `heartrate3`.  
```{r}
heartrate3 <- heartrate2 %>% 
  separate(patient, into=c("patient", "gender"), sep="_") %>% 
  pivot_longer(-c(patient, gender),
               names_to = "drug",
               values_to = "heartrate")
heartrate3
```
2. `unite()` is the opposite of separate(). Its syntax is straightforward. You only need to give a new column name and then list the columns to combine with a separation character.  Give it a try below by recombining patient and sex from `heartrate3`.  
```{r}
heartrate3 %>% 
  unite(patient_gender, "patient", "gender", sep="_")
```
## `pivot_wider()`
The opposite of `pivot_longer()`. You use `pivot_wider()` when you have an observation scattered across multiple rows. In the example below, `cases` and `population` represent variable names not observations.  
Rules:  
+ `pivot_wider`(names_from, values_from)  
+ `names_from` - Values in the `names_from` column will become new column names  
+ `values_from` - Cell values will be taken from the `values_from` column  
```{r}
tb_data <- read_csv("data/tb_data.csv")
tb_data
```

When using `pivot_wider()` we use `names_from` to identify the variables (new column names) and `values_from` to identify the values associated with the new columns.
```{r}
tb_data %>% 
  pivot_wider(names_from = "key",
              values_from = "value")
```
## Practice
1. Load the `gene_exp.csv` data as a new object `gene_exp`. Are these data tidy? Use `pivot_wider()` to tidy the data.
```{r}
gene_exp <- read_csv("data/gene_exp.csv")
gene_exp
```

```{r}
gene_exp %>% 
  pivot_wider(names_from = "type",
              values_from = "L4_values")
```
## Practice
For the last practice example, I will use data from the [R Ladies Sydney](https://rladiessydney.org/courses/ryouwithme/02-cleanitup-5/) blog. This data is compiled by the NSW Office of Environment and contains the _Enterococci_ counts in water samples obtained from Sydney beaches as part of the Beachwatch Water Quality Program.  

1. Load the beachbugs data and have a look.
```{r}
beachbugs <- read_csv("data/beachbugs_long.csv")
beachbugs
```
2. Use `pivot_wider` to transform the data into wide format.
```{r}
beachbugs_wide <- beachbugs %>% 
  pivot_wider(names_from = "site",
              values_from = "buglevels")
```
3. Now, use `pivot_longer` to transform them back to long!
```{r}
beachbugs_wide %>% 
  pivot_longer(-year,
               names_to="site",
               values_to = "buglevels")
```
## Tidy data
Most "wild" data are organized incorrectly for work in R and, as you might expect, the tools used to transform data are a core part of the tidyverse. I will attempt to summarize the most important points below, but you should read [chapter 5 of the data science text](https://r4ds.hadley.nz/data-tidy.html) for a more thorough explanation.  

`Tidy` data follows three conventions:   
(1) each variable has its own column  
(2) each observation has its own row  
(3) each value has its own cell  

## `pivot_longer()`  
Scientists use spreadsheets that are organized to make data entry efficient. This is often referred to as **wide format**. Unfortunately, the wide format creates a problem because column names may actually represent values of a variable. The command `pivot_longer()` shifts data from wide to long format.   

Rules:  
+ `pivot_longer`(cols, names_to, values_to)
+ `cols` - Columns to pivot to longer format
+ `names_to` - Name of the new column; it will contain the column names of gathered columns as values
+ `values_to` - Name of the new column; it will contain the data stored in the values of gathered columns

## Example 1: column names are data
The following data show results from a drug trial with four treatments on six patients. The values represent resting heart rate.
```{r}
heartrate <- read_csv("data/heartrate.csv")
heartrate
```

Let's assess whether or not these data are tidy.  
(1) each variable has its own column?  
 
(2) each observation has its own row?  

(3) each value has its own cell?  

To fix this problem, we need to reshape the table to long format while keeping track of column names and values. We do this using `pivot_longer()`. Notice that the dimensions of the data frame change.  
```{r}
heartrate_long <- heartrate %>% 
  pivot_longer(-patient,
               names_to= "drug",
               values_to = "heartrate")
```

```{r}
heartrate_long %>% 
  filter(patient=="Margaret") %>% 
  ggplot(aes(x=drug, y=heartrate))+
  geom_col()
```
## Practice
1. Import the file `relig_income.csv` and store it as a new object `relig_income`.  
```{r}
relig_income <- read_csv("data/relig_income.csv")
relig_income
```
2. Why are these data untidy? 
_Variable names are actually data._

3. Use `pivot_longer()` to make the data tidy.  
```{r}
relig_income %>% 
  pivot_longer(-religion, #don't move
               names_to = "income", #give new column name
               values_to = "total") #move the count into a new column
```
## Example 2: some column names are data
Some (but not all) of the column names are data. We also have NA's.
```{r}
billboard <- read_csv("data/billboard.csv")
billboard
```

Solution 1: specify a range of columns that you want to pivot.
```{r}
billboard %>% 
  pivot_longer(wk1:wk76,
               names_to="week",
               values_to = "rank",
               values_drop_na = T)
```
Solution 2: OR, specify columns that you want to stay fixed.
```{r}
billboard %>% 
  pivot_longer(-c(artist, track, date.entered),
               names_to = "week",
               values_to = "rank",
               values_drop_na = T)
```
Solution 3: identify columns by a prefix, remove the prefix and all NA's.
```{r}
billboard %>% 
  pivot_longer(
    cols=starts_with("wk"),
    names_to = "week",
    names_prefix = "wk", #gets rid of the wk
    values_to = "rank",
    values_drop_na = T
  )
```
## Practice  
1. Import `plant_data.csv` as a new object `plant_data`.  
```{r}
plant_data <- read_csv("data/plant_data.csv")
plant_data
```
2. Why are these data not tidy?  
_variable names are actually data_

3. Use `pivot_longer()` to make the data tidy. Focus the data only on genotype, water_sched_prog, and greenhouse.  
```{r}
plant_data %>% 
  pivot_longer(-c(genotype, water_sched_prog, greenhouse),
               names_to="v1",
               values_to = "v2",
               values_drop_na = T)
```
## Example 3: more than one variable in a column name
In this case, there are two observations in each column name.
```{r}
qpcr_untidy <- read_csv("data/qpcr_untidy.csv")
qpcr_untidy
```
`names_sep` helps pull these apart, but we still have "exp" and "rep" to deal with.  
```{r}
qpcr_tidy <- qpcr_untidy %>% 
  pivot_longer(
    exp1_rep1:exp3_rep3,
    names_to = c("experiment", "replicate"),
    names_sep = "_",
    values_to = "mRNA expression"
  )
```

```{r}
qpcr_tidy %>% 
  separate(experiment, into=c("experiment", "e_number"), sep=3) %>% 
  separate(replicate, into=c("rep", "r_number"), sep=3) %>% 
  select(gene, experiment=e_number, replicate=r_number, `mRNA expression`)
```
## Review
`summarize()` is a verb in `dplyr` that allows us to produce summaries of data. It's output is a new dataframe. In lab 7, we learned how `summarize()` can be used to produce quick statistical summaries.  

1. What is the average age, shoulder height, tusk length, and tusk circumference for all elephants in the data set?
```{r}
elephants %>% 
  summarize(across(where(is.numeric),
                   ~mean(.x, na.rm=T)))
```
We can include multiple functions using this same structure, but it becomes cumbersome.
```{r}
elephants %>% 
  summarize(
    across(
      where(is.numeric),
      list(
        mean   = ~mean(.x, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE),
        sd     = ~sd(.x, na.rm = TRUE),
        min    = ~min(.x, na.rm = TRUE),
        max    = ~max(.x, na.rm = TRUE),
        n      = ~sum(!is.na(.x))
      )
    )
  )
```
## `group_by()`
The `summarize()` function is often more useful when used with `group_by()`. `group_by()` allows us to group data by a categorical variable(s) of interest.   
```{r}
elephants %>% 
  group_by(sex)
```
Notice that the output doesn't look different, but two groups are indicated. Any verbs that follow `group_by()` will be applied to each group separately. For example, what is the average age, shoulder height, tusk length, and tusk circumference for males and females?
```{r}
elephants %>% 
  group_by(sex) %>% 
  summarize(across(where(is.numeric),
                   ~mean(.x, na.rm=T)))
```
We can group by more than one variable.
```{r}
elephants %>% 
  group_by(years_of_sample_collection, sex) %>% 
  summarize(mean_estimated_age_years=mean(estimated_age_years, na.rm=T),
            n=n(),
            .groups='keep') %>% 
  arrange(sex)
```
Do you remember this question from homework 7?  

"What is the mean, median, and standard deviation for age of males and females included in the study? Separate the results by year of sample collection. Does the sampling look even between years and sexes?"

Without `group_by()`, we had to produce multiple code chunks- we are now more efficient!
```{r}
elephants %>% 
  filter(years_of_sample_collection=="1966-68") %>% 
  group_by(sex) %>% 
  summarize(mean_age=mean(estimated_age_years, na.rm=T),
            median_age=median(estimated_age_years, na.rm=T),
            sd_age=sd(estimated_age_years, na.rm=T),
            n=n())
```

```{r}
elephants %>% 
  filter(years_of_sample_collection=="2005-13") %>% 
  group_by(sex) %>% 
  summarize(mean_age=mean(estimated_age_years, na.rm=T),
            median_age=median(estimated_age_years, na.rm=T),
            sd_age=sd(estimated_age_years, na.rm=T),
            n=n())
```

```{r}
elephants %>% 
  group_by(sex, years_of_sample_collection) %>% 
  summarize(mean_age=mean(estimated_age_years, na.rm=T),
            median_age=median(estimated_age_years, na.rm=T),
            sd_age=sd(estimated_age_years, na.rm=T),
            n=n(),
            .groups='keep')
```
## Practice
Use the built-in `msleep` data to answer the following questions.  

1. Calculate the min, max, mean, and total number of observations for body weight by vore.  
```{r}
names(msleep)
```

```{r}
msleep %>% 
  group_by(vore) %>% 
  summarize(min_bodywt=min(bodywt),
            max_bodywt=max(bodywt),
            mean_bodywt=mean(bodywt),
            n=n())
```
2. Calculate mean brain weight by taxonomic order in the msleep data.
```{r}
msleep %>% 
  group_by(order) %>% 
  summarize(mean_brainwt=mean(brainwt))
```
3. Try running the code again, but this time add `na.rm=TRUE`. What is the problem with Cetacea?
```{r}
msleep %>% 
  group_by(order) %>% 
  summarize(mean_brainwt=mean(brainwt, na.rm=T))
```

```{r}
msleep %>% 
  filter(order=="Cetacea") %>% 
  select(order, genus, brainwt)
```
## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```
## Data
For this lab, we will use the built-in data on mammal sleep patterns, `msleep`. From: _V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007_.

```{r}
msleep
```
## Summary statistics
When first exploring data, it is often a good idea to produce some basic summary statistics. This can help you identify potential problems with the data (e.g., outliers, missing values, skew) as well as get a general sense of the distribution of values for a given variable. So far, you have learned to calculate summary statistics using _base R_ functions such as `mean()`, `sd()`, `min()`, and `max()`.  
```{r}
summary(msleep)
```
## Practice
1. Calculate the mean body weight of large and small mammals in msleep. Large mammals have a body weight greater than 150, while small mammals have a body weight less than 10. I did large mammals for you as an example.
```{r}
large <- msleep %>% 
  filter(bodywt > 150) %>% 
  arrange(desc(bodywt))
large
```

```{r}
small <- msleep %>% 
  filter(bodywt<10) %>% 
  arrange(desc(bodywt))
small
```

```{r}
mean(large$bodywt)
```

```{r}
mean(small$bodywt)
```
## `summarize()`
Fortunately, dplyr has some functions that make producing summary statistics easier. The verb `summarize()` produces summary statistics for a given variable in a data frame. Let's try the examples above using `summarize()`. 

Large mammals
```{r}
msleep %>% 
  filter(bodywt>150) %>% 
  summarize(mean_bodywt_lg=mean(bodywt))
```
Small mammals
```{r}
msleep %>% 
  filter(bodywt<10) %>% 
  summarize(mean_bodywt_sm=mean(bodywt))
```
You can also combine functions to make summaries for multiple variables.
```{r}
msleep %>% 
  filter(bodywt>150) %>% 
  summarize(mean_bodywt_lg=mean(bodywt),
            min_bodywt_lg=min(bodywt),
            max_bodywt_lg=max(bodywt),
            sd_bodywt_lg=sd(bodywt),
            total=n())
```
Maybe you want to summarize all of the numeric variables in the data frame. You can use the `where()` function to select only the numeric variables.
```{r}
msleep %>% 
  select(where(is.numeric)) %>% 
  summarize_all(mean, na.rm=T)
```
This is a little more efficient.
```{r}
msleep %>% 
  summarize(across(where(is.numeric),
                   ~mean(.x, na.rm=T)))
```
## Practice
1. What is the mean, min, and max `bodywt` for the taxonomic order Primates? Provide the total number of observations.
```{r}
msleep %>% 
  filter(order=="Primates") %>% 
  summarize(mean_bodywt=mean(bodywt),
            min_bodywt=min(bodywt),
            max_bodywt=max(bodywt))
```
`n_distinct()` is a handy way of cleanly presenting the number of distinct observations. Notice that there are 9 genera with over 100 in body weight.
```{r}
msleep %>% 
  filter(bodywt>100) %>% 
  distinct()
```
Here we show the number of distinct genera over 100 in body weight.
```{r}
msleep %>% 
  filter(bodywt>100) %>% 
  summarize(n_genera=n_distinct(genus))
```
Not to be confused with n(), which counts the total number of observations.
```{r}
msleep %>% 
  filter(bodywt>100) %>% 
  summarize(total=n())
```
`pull()` is another useful verb that extracts a single column from a data frame as a vector.
```{r}
msleep %>% 
  filter(bodywt>100) %>% 
  distinct(genus) %>% 
  pull(genus)
```
## Practice
1. How many genera are represented in the msleep data frame?
```{r}
msleep %>% 
  summarize(n_genera=n_distinct(genus))
```
## Counts
Although `summarize()` is helpful, oftentimes we are only interested in counts. The [janitor package](https://garthtarr.github.io/meatR/janitor.html) does a lot with counts, but there are also functions that are part of dplyr that are useful.  

For example, if we wanted to count the number of observations for each `vore` category in the `msleep` data frame we could do the following:
```{r}
msleep %>% 
  count(vore, sort=T)
```

```{r}
msleep %>% 
  count(vore) %>% 
  ggplot(aes(x=vore, y=n))+
  geom_col() #use this when we have an x and a y
```
You can even combine multiple variables to get counts. For example, if we wanted to count the number of observations for each `vore` category within each `order` category we could do the following:
```{r}
msleep %>% 
  count(vore, order)
```
The tabyl() function from the janitor package is also very useful for producing counts and percentages.  
```{r}
msleep %>% 
  tabyl(vore)
```
## Practice
1. In the taxonomic order Carnivora, count the number of observations for each conservation status.
```{r}
msleep %>% 
  filter(order=="Carnivora") %>% 
  count(conservation, sort=T)
```
2. Among herbivores, which order is most represented? Return the top 5 orders with counts, sorted in descending order.
```{r}
msleep %>% 
  filter(vore=="herbi") %>% 
  count(order, sort=T)
```
1. Select species, island, and body mass from the penguins data. Arrange results by body mass.
```{r}
penguins %>% 
  select(species, island, body_mass_g) %>% 
  arrange(body_mass_g)
```
2. Filter the penguins data to only include observations from Biscoe and Dream islands.
```{r}
penguins %>% 
  filter(island=="Biscoe" | island=="Dream")
```

```{r}
penguins %>% 
  filter(island==c("Biscoe", "Dream"))
```

```{r}
penguins %>% 
  filter(island %in% c("Biscoe", "Dream"))
```
3. Make a plot that shows the relationship between body mass and flipper length. How does this compare among different species?
```{r}
penguins %>% 
  ggplot(aes(x=flipper_length_mm, y=body_mass_g, color=species)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)
```
## `distinct()`
`distinct()` looks for all unique observations in rows. This is a little tricky because it can look like it is working column-wise, but it is actually working row-wise.  

One helpful approach to new data is to find any duplicated rows. If we first look at the dimensions of the penguins data, we see it has 344 rows and 8 columns.
```{r}
dim(penguins)
```
Using `distinct()` across all rows, we see there are no duplicates. This means every row contains unique observations across all variables.
```{r}
penguins %>% 
  distinct()
```
But if we only look at species, we can see that there are only 3 unique species in the data.
```{r}
penguins %>% 
  distinct(species)
```
What if we want to know which islands each species occurs on?
```{r}
penguins %>% 
  distinct(species, island, .keep_all=TRUE)
```
## `mutate()`
`mutate()` is another verb that acts on columns. It allows us to create new columns from existing columns in a data frame. When we use `mutate()`, the columns are added to the end of the dataframe by default. Let's create a new column that converts body mass from grams to kilograms.
```{r}
penguins %>%
  mutate(body_mass_kg = body_mass_g/1000) %>% 
  select(species, body_mass_g, body_mass_kg) %>% 
  arrange(body_mass_kg)
```
## `mutate()` and `across()`
We use `across()` within `mutate()` to apply a function to multiple columns. This is especially helpful when cleaning data. For example, let's say we want to convert all columns that end with `mm` to centimeters. We can use `across()` to do this.
```{r}
penguins %>%
  mutate(across(ends_with("mm"), ~./10)) %>%
  select(species, 
         bill_length_cm=bill_length_mm, 
         bill_depth_cm=bill_depth_mm, 
         flipper_length_cm=flipper_length_mm)
```
What does the `~./10` mean? The `~` indicates that what follows is a formula (lambda function). The `.` represents the current column being processed. So, `./10` means "take the current column and divide it by 10". This operation is applied to all columns that end with `mm`.

## Cleaning Data
```{r}
mammals <- read_csv("data/mammal_lifehistories_v2.csv")
```
What is the structure of the data? Are there any NA's or other issues?
```{r}
glimpse(mammals)
```
One thing to notice is the column names are inconsistent. This is going to cause problems for us down the line. We could rename each column, one at a time, using `rename()`, but that would be tedious. Instead, we can use the `clean_names()` function from the `janitor` package to fix all of the column names at once.
```{r}
mammals <- mammals %>% 
  clean_names()
```

```{r}
glimpse(mammals)
```
Notice that `clean_names()` has converted all column names to lowercase and replaced spaces with underscores. But, no adjustments were made to the data itself. What if we want to change observations from upper case to lower case?
```{r}
mammals %>% 
  mutate(across(c("order", "family"), tolower)) #specific columns
```
This will change all columns to lower case. But, notice what happens to numeric columns.
```{r}
mammals %>% 
  mutate(across(everything(), tolower)) #all columns
```
For this reason, it might be better to use `where` so we can specify only character columns.
```{r}
mammals <- mammals %>%
  mutate(across(where(is.character), tolower)) #all character columns
```
## `if_else()`
We briefly introduce `if_else()` here because it allows us to use `mutate()` but not have the entire column affected in the same way. With `ifelse()`, you first specify a logical statement, afterwards what needs to happen if the statement returns `TRUE`, and lastly what needs to happen if it's  `FALSE`.  

Have a look at the data from mammals below. Notice that the values for newborn include `-999.00`. This is sometimes used as a placeholder for NA (but, is a really bad idea). We can use `if_else()` to replace `-999.00` with `NA`.  
```{r}
mammals %>% 
  select(genus, species, newborn) %>% 
  arrange(newborn)
```

```{r}
mammals %>% 
  select(genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))%>% 
  arrange(newborn)
```
## Practice
1. Following the example above, convert all -999 values in the mammals dataframe to NA.
```{r}
mammals <- mammals %>%
  mutate(across(c(mass, wean_mass, gestation, max_life, newborn, weaning, litter_size, afr, litters_year),
                ~ifelse(. == -999, NA, .)))
```

```{r}
summary(mammals)
```
2. In the mammals data, make a new column `mass_kg` that that converts mass from grams to kilograms. Select the columns genus, species, mass, and mass_kg, and arrange the data by mass_kg in descending order. What is the common name for the species with the highest mass?  
_blue whale_
```{r}
mammals %>% 
  mutate(mass_kg = mass/1000) %>% 
  select(genus, species, mass, mass_kg) %>% 
  arrange(desc(mass_kg))
```
3. What is the relationship between gestation and newborn mass?
```{r}
mammals %>% 
  mutate(mass_kg = mass/1000) %>% 
  mutate(wean_gestation_ratio = log10(newborn/gestation)) %>% 
  select(genus, species, wean_gestation_ratio) %>% 
  arrange(desc(wean_gestation_ratio))
```

```{r}
mammals %>% 
  ggplot(aes(x=gestation, y=log10(newborn))) +
  geom_point()+
  geom_smooth(method="lm", se=FALSE)
```
4. Which mammal has the longest life span in years?  
_fin whale_
```{r}
mammals %>% 
  select(family, genus, species, max_life) %>% 
  mutate(max_life_new = max_life/12) %>%
  arrange(desc(max_life_new))
```
5. Do larger mammals live longer?
```{r}
mammals %>% 
  ggplot(aes(x=log10(mass), y=max_life)) +
  geom_point()+
  geom_smooth(method="lm", se=FALSE)
```
