---
title: "Homework 3"
author: "Andrei Graham"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run.  

## Load the tidyverse
```{r}
library("tidyverse")
```

In order to answer the questions below, you will need to do a little online research.

**1. Make three new vectors that show the name, height in feet, and height in meters of the five tallest mountains in the world.**
```{r}
#mount_everest 8849,29032
#aconcagua 6962,22829
#denali 6194,20320
#kilimanjaro 5895,19340
#mount_elbrus 5642,18510

meters<-c(8849,6962,6194,5895,5642)
feet<-c(29032,22829,20320,19340,18510)
mt_name<-c("everest","aconcagua","denali","kilimanjaro","elbrus")

```

**2. Combine these vectors into a data frame called `mountains`.**
```{r}
mountains<-tibble(meters,feet,mt_name)
```

**3. What is the mean height of the mountains in feet?**
```{r}
mean(mountains$feet)
```

**4. When were each of these mountains first climbed (i.e. in what year)? Make a new vector `first_climbed` and add it to the `mountains` data frame.**
```{r}
first_climbed<-c(1953,1897,1913,1889,1829)
mountains<-tibble(first_climbed,feet,meters,mt_name)
mountains
```

```{r}

```

**5. How many times have each of these mountains been climbed? Make a new vector `summits` and add it to the `mountains` data frame.**
```{r}
#everest has been climbed 12884 times
#denali has been climbed 26358 times
#aconcagua NA
#elbrus NA
#kilimanjaro NA

summits<-c(12884,26358,NA,NA,NA)
mountains<-tibble(first_climbed,feet,meters,mt_name,summits)
```

**6. Which mountain has the highest number of fatalities? Make a new vector `fatalities` and add it to the `mountains` data frame.**
```{r}
#everest has around 330 deaths
#denali has around 130 deaths
#aconcagua has around 160 deaths
#elbrus is unknown
#kilimanjaro is unknown

#everest has the highest number of fatalities out of the mountains

fatalities<-c(330,130,160,NA,NA)

mountains<-tibble(first_climbed,feet,meters,mt_name,summits,fatalities)
```

**7. What is the fatality rate (i.e., fatalities divided by summits) for each mountain? Create a new vector `fatality_rate` and add it to the `mountains` data frame.**
```{r}
na.omit(fatalities)
na.omit(summits)
(fatalities/summits)

fatality_rate<-c(fatalities/summits)

mountains<-tibble(first_climbed,feet,meters,mt_name,summits,fatalities,fatality_rate)
```

**8. Write your data frame to a .csv file called `mountains_data`.**
```{r}
write.csv(mountains,file = "mountains_data.csv",row.names=FALSE )
```

**9. Clear your environment panel by clicking on the broom icon. Then read in your `mountains_data.csv` file to a new object called `mountains`.**
```{r}
mountains<-read.csv("mountains_data.csv")
```

```{r}

```

**10. Use a summary function of your choice to show the structure of your `mountains` data frame.**
```{r}
summarise(mountains)
```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  